Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-08-08 latex/intro.tex
Add explanation following pazman pronzato

It is important to note that we do not view measurement clusterization
as undesirable, nor do we believe it should be avoided at
all. Clusterization is a peculiar phenomenon and it is perfectly
reasonable for someone to argue against the D-optimality criterion
based on the fact that it results in clustered designs. We have seen,
however that there is a perfectly reasonable explanation for
clusterization. We have shown that clusterization is an inevitable
consequence of having a problem with some modes where uncertainty
decays faster than others.

@@

We believe that the emergence of clusterization in this context is
thus non-physical, arising from the way the inverse problems we
consider are phrased. Clusterization therefore indicates that the
underlying mathematical model is overly permissive and fails to
capture crucial physical constraints of the problem. We suggest that
when clusterization occurs, practitioners should consider alternative
models where information is localized in space and travels at finite
speed in the medium. Such models may not only provide more physically
accurate and meaningful results but may also mitigate the issue of
clusterization.

>>>>

It is important to note that we do not view clustered designs as
undesirable, nor do we believe it should be avoided at
all. Clusterization is a peculiar phenomenon and it is perfectly
reasonable for someone to argue against the D-optimality criterion
based on the fact that it results in clustered designs. We have seen,
however that there is a perfectly reasonable explanation for
clusterization. We have shown that clusterization is an inevitable
consequence of having a problem with some modes where uncertainty
decays faster than others.

@@

Our choice of prior is also a potential major contributor to
clusterization. Our choice of Gaussian prior similarly implies
information is shared between distant locations in $\Omega$. Thus, we
suggest refraining from choosing Gaussian priors with inverse
Laplacian covariance operators. Rather, non-Gaussian priors could be
employed instead \cite{hosseini2017, hosseini2019}.

We believe that the emergence of clusterization in this context is
thus non-physical, arising from the way the inverse problems we
consider are phrased. Clusterization therefore indicates that the
underlying mathematical / Bayesian model is overly permissive and
fails to capture crucial physical constraints of the problem. We
suggest that when clusterization occurs, practitioners should consider
alternative models where information is localized in space and travels
at finite speed in the medium. Such models may not only provide more
physically accurate and meaningful results but may also mitigate the
issue of clusterization.

### 2024-08-08 latex/numerix.tex
Add explanation following pazman pronzato

To simplify things, we directly sample rank $\opt^*\opt$. We iterate
over the number of measurements $m \in \{4,\dots, 24\}$, and for every
$m$ we then iterate over $k:=\rank \obs^*\obs \in \{2,\dots,
m-1\}$. For each pair $m,k$ we repeat the following steps $N=5000$
times:
\begin{enumerate}
\item Generate random diagonal $D\in \mathbb{R}^{k\times k}$ with
  entries $\log (d_i) \sim \mathcal{N}(50,15)$ and normalize so that
  $\ttr D = m$\footnote{Results are robust to the choice of parameters
  in the lognormal distribution above.}.
\item Conjugate $D$ by a random orthogonal matrix to form a positive
  semi-definite $M := UDU^t \in \mathbb{R}^{k\times k}$. This $M$
  represents $\opt^*\opt$.
\item Apply the construction of Lemma \ref{lemma:free} to calculate
  $A$ such that $AA^t = M$, where $A$ has unit norm columns. I.e.~find
  the optimal design $\opt$.
\item Since $A$ corresponds to $\opt$, its columns correspond to
  measurement vectors. We call $A$ "clustered" if $A$ has two or more
  identical columns (up to some numerical precision threshold).
\end{enumerate}
We then calculate the fraction of clustered designs of the simulations
we ran, for each pair $m,k$. Clusterization occurred at high rates
($>99.9\%$) whenever $m-k > 1$; see Fig.~\ref{fig:sim_AAt}. Hence, in
these simulations, clusterization is a generic property. However, when
$m-k = 1$, clusterization does not occur. We do not why this is so and
further investigation into this phenomenon is out of scope for the
current study.

>>>>

To simplify things, we directly sample rank $\opt^*\opt$. We iterate
over the number of measurements $m \in \{4,\dots, 24\}$, and for every
$m$ we then iterate over $k:=\rank \obs^*\obs \in \{2,\dots,
m-1\}$. For each pair $m,k$ we repeat the following steps $N=5000$
times:
\begin{enumerate}
\item Generate random diagonal $D\in \mathbb{R}^{k\times k}$ with
  entries $\log (d_i) \sim \mathcal{N}(50,15)$ and normalize so that
  $\ttr D = m$. 
\item Conjugate $D$ by a random orthogonal matrix to form a positive
  semi-definite $M := UDU^t \in \mathbb{R}^{k\times k}$. This $M$
  represents $\opt^*\opt$.
\item Apply the construction of Lemma \ref{lemma:free} to calculate
  $A$ such that $AA^t = M$, where $A$ has unit norm columns. I.e.~find
  the optimal design $\opt$.
\item Since $A$ corresponds to $\opt$, its columns correspond to
  measurement vectors. We call $A$ "clustered" if $A$ has two or more
  identical columns (up to some numerical precision threshold).
\end{enumerate}
We then calculate the fraction of clustered designs of the simulations
we ran, for each pair $m,k$. Clusterization occurred at high rates
($>99.9\%$) whenever $m-k > 1$; see Fig.~\ref{fig:sim_AAt}. Hence, in
these simulations, clusterization is a generic property. However, when
$m-k = 1$, clusterization does not occur. We do not why this is so and
further investigation into this phenomenon is out of scope for the
current study.

### 2024-08-08 latex/vanishing.tex
Text ready, answer refs
Add explanation following pazman pronzato

\subsection{Answer to Question \ref{q:why}}\label{subsec:why}
Building on Theorem \ref{thm:char}, we can now give a compelling
explanation to the measurement clusterization we observed for the
inverse problem of the heat equation (Fig.~\ref{fig:eigenvectors}).

Consider $\fwd$ and $\prcov$ from \emph{the inverse problem of the
heat equation}. As before, we denote the eigenvalues of
$\fwd\prcov\fwd^*$ by $\lambda_j$. We input these eigenvalues into our
\emph{generic} model, and find a D-optimal design $\opt$ for our
generic model using Theorem \ref{thm:char}. In our generic model, the
measurements we take are best utilized in reducing uncertainty for the
first $k$ eigenvectors. So, a D-optimal design arising from our
\emph{generic model} completely avoids measuring eigenvectors $k+1$
and above.

@@

The abovementioned trade-off is explored in
Fig.~\ref{fig:eigenvectors}. We allow $m=4$ measurements in $\Omega =
[0,1]$ and observe that D-optimal measurement locations are clustered
at $x_1 = 0.31$ and $x_2 = 0.69$. Upon close inspection of the scaled
eigenvectors of $\fwd \prcov \fwd^*$, we first observe that
eigenvectors $3$ and above have negligible prior amplitude. Since we
only have $m=4$ measurements at our disposal, we interpret these
results, following Theorem \ref{thm:char}, as implying we should only
care about measuring the first and second eigenvectors. Then, we note
the D-optimal $x_1,x_2$ present a compromise between the amplitude of
the first and second eigenvectors. For example, a measurement at
$x=0.5$ would have ignored the second eigenvector altogether, since
the second eigenvector is zero at $x=0.5$.

@@

>>>>

\subsection{Answer to Question \ref{q:why}}\label{subsec:why}
Building on Theorem \ref{thm:char}, we can now give a compelling
explanation to the measurement clusterization we observed for the
inverse problem of the heat equation, see Section
\ref{subsub:clusterization1} below. We also suggest a generic
explanation for clusterization, see Section \ref{subsub:clusterization2}.

\subsubsection{Clusterization in the heat equation}
Consider $\fwd$ and $\prcov$ from \emph{the inverse problem of the
heat equation}. As before, we denote the eigenvalues of
$\fwd\prcov\fwd^*$ by $\lambda_j$. We input these eigenvalues into our
\emph{generic} model, and find a D-optimal design $\opt$ for our
generic model using Theorem \ref{thm:char}. In our generic model, the
measurements we take are best utilized in reducing uncertainty for the
first $k$ eigenvectors. So, a D-optimal design arising from our
\emph{generic model} completely avoids measuring eigenvectors $k+1$
and above.

@@

The abovementioned balance is explored in
Fig.~\ref{fig:eigenvectors}. We allow $m=4$ measurements in $\Omega =
[0,1]$ and observe that D-optimal measurement locations are clustered
at $x_1 = 0.31$ and $x_2 = 0.69$. Upon close inspection of the scaled
eigenvectors of $\fwd \prcov \fwd^*$, we first observe that
eigenvectors $3$ and above have negligible prior amplitude. Since we
only have $m=4$ measurements at our disposal, we interpret these
results, following Theorem \ref{thm:char}, as implying we should only
care about measuring the first and second eigenvectors. Then, we note
the D-optimal $x_1,x_2$ present a compromise between the amplitude of
the first and second eigenvectors. For example, a measurement at
$x=0.5$ would have ignored the second eigenvector altogether, since
the second eigenvector is zero at $x=0.5$.

@@

\subsubsection{Clusterization in our generic model}
In our generic model, the limit on the number of measurements arises
as a consequence of Carath\'eodory's Theorem\footnote{Let $X \subseteq
\mathbb{R}^n, X \neq \phi$ and denote $\conv (X)$ the convex hull of
$X$. Let $x \in \conv (X)$. Then $x$ is a convex combination of at
most $n+1$ vectors in $X$.} and the concentration on the first $k$
eigenvectors of $\fwd \prcov \fwd^*$. The discussion below extends
arguments presented in \cite[Chapter 3]{silvey} and \cite[Section
  5.2.3]{pronzatoPazman2013}.

We now ignore all but the first $k$ eigenvectors of $\fwd \prcov
\fwd$, as instructed by Theorem \ref{thm:char}. Thus, we can safely
replace $\hilo$ with the $k$-dimensional subspace spanned by the first
$k$ eigenvectors of $\fwd^*\prcov\fwd$. Since this subspace is
isomorphic to $\mathbb{R}^k$, we do not name it and simply denote it
by $\mathbb{R}^k$. Now, let
\begin{equation*}
  \mathcal{M} := \conv \{\meas \meas^* : \meas\in \mathbb{R}^k, \|\meas\|=1\},
  %\hilo^{(k)}\},
\end{equation*}
where $\conv$ denotes the convex hull of a set. Since $\mathcal{M}$ is
a set of positive-definite matrices, it lives in a
$k(k+1)/2$-dimensional subspace of $\mathbb{R}^{k\times k}$. Now note
that $\frac{1}{m}\opt^*\opt \in \mathcal{M}$. By Carath\'eodory's
Theorem, there exists a convex combination
\begin{equation*}
  \frac{1}{m} \opt^*\opt = \sum_{i=1}^{\frac{k(k+1)}{2} + 1} \alpha_i \meas\meas^*,
\end{equation*}
where $\|\meas\|=1, \sum_{i=1}^I\alpha_i = 1$ and $\alpha_i > 0$.

For example, if $k=2$ --- as is the case for the inverse prblem of the
heat equation --- then an optimal design $\opt$ is a combination of at
most $3$ measurement vectors $\meas$. Thus, if the number of allowed
measurements $m$ is larger than $3$, then there exists an optimal
clustered design.

%%  whereThus, if we view $\obs$ as a $\mathbb{R}^{m\times k}$ matrix, we
%%  conclude that $\obs^*\obs$ is a symmetric positive definite matrix
%%  that lives in a $k(k+1)/2$-dimensional space.

%% The set $\conv \mathcal{M}$ is a superset of all possible

%% Since we restrict $\|\meas\| = 1$, the set of allowed
%% $\obs^*\obs$ is the ball of norm $m$ unit ball in (a space isomorphic to)
%% $\mathbb{R}^k.

%% $\{\meas \meas: \meas \in \span (\ev_1,\ev_2,\dots,
%% \ev_k\}$ is in fact $k$-dimensional.

%%   By Theorem
%%   \ref{thm:char}, we ignore all eigenvectors except for the first
%%   $k$.\ \

%% First, note that $\fwd^*\obs^*\obs\fwd$ is the Fisher inf
